### 2020-03-03 rebuttal/rebuttal.tex

\begin{figure}[h!]
	\centering
	\includegraphics{Fig10_bifurcation.eps}
	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
		The blue line shows the positions of the deterministic fixed points, while the dashed line indicates the instability of the middle one. The orange and yellow points show the mean of the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. To make this clearer, the bimodal distributions themselves are shown in a light color (in a smoothed form). The arrows illustrate the hysteresis loop in the stochastic simulations.}
	\label{bifucation}
\end{figure}

\begin{review}
	(and seems to take an unreasonably long time to switch between states).
\end{review}

\begin{response}
To address the indeed valid point of unreasonable switching times, we
have compared the timescales in our stochastic simulations with those from the
numerical solution of the deterministic model. We find excellent
agreement, as shown in an updated figure in the main text and also below
in Fig~\ref{adaptationtime}. So, again perhaps surprisingly,
stochasticity does not materially alter the mean switching time as
computed from the deterministic picture. Even adding additional
stochasticity through transcriptional bursting (discussed more below)
does not alter the mean switching time, though it does broaden the
distribution of times.
It is possible that this broadening resolves the apparent paradox. Even
if the mean switching time is quite long, a non-negligible fraction of a
population of cells could nevertheless switch on a shorter time-scale,
especially with inclusion of extra stochasticity beyond what we have
explicitly considered.
\end{response}

	\begin{figure}[h!]
		\centering
		\includegraphics{Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Blue shows the results from one typical run of the stochastic simulation, light blue shows two more extreme runs, and orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was started
			at an mRNA, protein and intracellular xanthosine count of 0. The
			extracellular xanthosine concentration was chosen to be
			$\mathrm{[c]_a} = 25$. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The orange line shows the corresponding deterministic time. The green dashed line right next to it indicated the mean of the blue distribution.}
		\label{adaptationtime}
	\end{figure}

### 2020-03-03 main_text/main_text.tex
More of GC's comments
Finished main_text GC comments

	To elaborate on the preceeding statements,
	we found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger distance between the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.

	Note that while the deterministic analysis assumes the variables to be
	continuous, the simulations work with discrete numbers of mRNA, protein, and
	xanthosine. This per se is no problem, because the deterministic analysis
	describes the mean values and the simulation fluctuates around this mean.
	If, however, the mRNA number at the third (high) fixed point is too low, the
	system may not be able to resolve the two points anymore. The tolerance to this is surprisingly large, though: In Fig~\ref{fig8:stochC}, the distance between the first and the third fixed point is around 3 mRNA molecules. While this is around the size of the fluctuations in Fig~\ref{fig8:stochC}, the latter become as large as 10 mRNA molecules when bursts are included in the simulation (see~\nameref{S1_Text}) and yet, the system is able to resolve the fixed points very well.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Shown in blue is the result from one run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}.
			(B) In blue we plot the waiting time distribution for 1000
			runs of the simulation (same conditions as (A)) to reach
			the steady-state mean.
			We define this time as the elapsed time when the protein
			copy number first reaches 90\% of its value at the upper
			fixed point.
			To better visualize the bulk of the distribution, we
			excluded from the plot $\sim10$ runs with adaptation times
			larger than $6 \cdot 10^5 \unit{s}$.
			The green dashed line indicates the mean of the blue distribution.
			The orange line shows the corresponding deterministic time.
			}
		\label{fig9:StochT}
	\end{figure}

@@

	\section*{Conclusion}
	In this paper, we propose a simple model for genetic circuits containing
	a membrane transporter whose gene expression is, directly or indirectly,
	activated by its substrate. We have shown that such a system can be bistable
	and thus work as a genetic switch which reacts to the extracellular
	concentration of the relevant metabolite. This switch has very useful
	biological features. First, coupling of the transporter with, for example, an
	enzyme which metabolizes the transporter substrate creates a genetic switch that enables short-term
	adaptation of the cell's metabolism to its environment. Second, the switch
	is stabilized by hysteresis effects when the extracellular substrate
	concentration decreases, which explains previous experimental findings.
	Furthermore, our simulations show that the switch-like behavior is very robust.

@@

	Despite the small copy numbers at the lower fixed point, the stochastic simulations are in excellent agreement with deterministic predictions. All model parameters could be reasonably estimated despite the paucity of
	experimental knowledge about the model system. The concentrations of mRNA, protein, and xanthosine at the fixed points as well as all qualitative features are as expected from similar systems and the few experiments on the \emph{xap} circuit. These points suggest that the model captures the relevant components of the system correctly and is able to describe its
	dynamics. 
	Furthermore, the modeling results let us, to some extent, understand why the biological system is constructed the way it is. By keeping the model as minimal as possible, but still modeling every part explicitly with an appropriate complexity, we can investigate the interesting features while still being able to understand the influence of all parameters and their interplay intuitively.

@@

	\section*{Acknowledgments}
	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations.
	Portions of the experiments reported here were carried out at the
	Physiology Course at the Marine Biological Laboratory in Woods
	Hole, MA, operated by the University of Chicago.
	This work was supported by 1R35 GM118043-01 Maximizing Investigators'
	Research Award (MIRA) (to R.P.), and the Werner Siemens Foundation
	through the Swiss Study Foundation (K.S.L.). This material is based upon
	work supported by the National Science Foundation Graduate Research
	Fellowship under Grant No.\ DGE-1745301 (M.J.M.).

>>>>

	To elaborate on the preceeding statements,
	we found that the two lower fixed points (marked as 1 and 2 in
	Fig~\ref{fig4:bistability}) need to be very close like in
	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
	meaning larger difference between the concentrations at the first and second fixed point,
	almost no switching was observed. Of course, switching is also a matter of the
	waiting time and stochastic effects: if one waits for long enough, it
	should eventually occur. However, switching times of more than several hours
	are not at the center of this investigation and would mean that switching is
	extremely unlikely. There are two aspects that become relevant in this
	context that we neglect in our analysis but briefly mention here:
	transcription and translation bursts lead to higher stochasticity and cell
	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.

	Note that while the deterministic analysis assumes the variables to be
	continuous, the simulations work with discrete numbers of mRNA, protein, and
	xanthosine. This per se is no problem, because the deterministic analysis
	describes the mean values and the simulation fluctuates around this mean.
	However, if the mRNA number of the third (high) fixed point is so low that stochastic fluctuations are larger than the difference in concentrations between the first and the second fixed point, the
	system may not be able to resolve the two points anymore. The tolerance to this is surprisingly large, though: In Fig~\ref{fig8:stochC}, the distance between the first and the third fixed point is around 3 mRNA molecules. While this is around the size of the fluctuations in Fig~\ref{fig8:stochC}, the latter become as large as 10 mRNA molecules when bursts are included in the simulation (see~\nameref{S1_Text}) and yet, the system is able to resolve the fixed points very well.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Shown in blue is the result from one typical run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's. In light blue, two more extreme runs of the simulation are shown for comparison. The simulation was run for $5 \cdot 10^5 \unit{s}$ each and started
			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
			Table~\ref{table1:nondim}, the only exception being the
			extracellular xanthosine concentration, which was chosen to be
			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
			Fig~\subref{fig8:stochC}{(C)}.
			(B) In blue we plot the waiting time distribution for 1000
			runs of the simulation (same conditions as (A)) to reach
			the steady-state mean.
			We define this time as the elapsed time when the protein
			copy number first reaches 90\% of its value at the upper
			fixed point.
			To better visualize the bulk of the distribution, we
			excluded from the plot $\sim10$ runs with adaptation times
			larger than $6 \cdot 10^5 \unit{s}$.
			The green dashed line indicates the mean of the blue distribution.
			The orange line shows the corresponding deterministic time.
			}
		\label{fig9:StochT}
	\end{figure}

@@

	\section*{Conclusion}
	Here, we propose a simple model for genetic circuits containing
	a membrane transporter whose gene expression is, directly or indirectly,
	activated by its substrate. We have shown that such a system can be bistable
	and thus work as a genetic switch which reacts to the extracellular
	concentration of the relevant metabolite. This switch has very useful
	biological features. First, coupling of the transporter with, for example, an
	enzyme which metabolizes the transporter substrate creates a genetic switch that enables short-term
	adaptation of the cell's metabolism to its environment. Second, the switch
	is stabilized by hysteresis effects when the extracellular substrate
	concentration decreases, which explains previous experimental findings.
	Furthermore, our simulations show that the switch-like behavior is very robust.

@@

	Despite the small copy numbers at the lower fixed point, the stochastic simulations are in excellent agreement with deterministic predictions. All model parameters could be reasonably estimated despite the paucity of
	experimental knowledge about the model system. The concentrations of mRNA, protein, and xanthosine at the fixed points as well as all qualitative features are as expected from similar systems and the few experiments on the \emph{xap} circuit. These points suggest that the model captures the relevant components of the system correctly and is able to describe its
	dynamics. 
	The modeling results let us, to some extent, understand how the biological system can achieve its function. By keeping the model as minimal as possible, but still modeling every part explicitly with an appropriate complexity, we can investigate the interesting features while still being able to understand the influence of all parameters and their interplay intuitively.

@@

	\section*{Acknowledgments}
	We thank Jane Kondev and Jin Wang for interesting discussions, Patrick Lenggenhager for his help with \emph{Mathematica} issues, and Nigel Orme for assistance with illustrations.
	Portions of the experiments reported here were carried out at the
	Physiology Course at the Marine Biological Laboratory in Woods
	Hole, MA, operated by the University of Chicago.
	This work was supported by 1R35 GM118043-01 Maximizing Investigators'
	Research Award (MIRA) (to R.P.), and the Werner Siemens Foundation
	through the Swiss Study Foundation (K.S.L.). This material is based upon
	work supported by the National Science Foundation Graduate Research
	Fellowship under Grant No.\ DGE-1745301 (M.J.M.).
